Accelerating Neural Network Training using Arbitrary Precision Approximating Matrix Multiplication Algorithms

### 2021-04-22 latex/ICPP21/paper.tex

\documentclass[sigconf]{acmart}

@@

%\author{\IEEEauthorblockN{Grey Ballard}
%\IEEEauthorblockA{\textit{Department of Computer Science} \\
%\textit{Wake Forest University}\\
%Winston-Salem, NC, USA \\
%\email{ballard@wfu.edu}}
%\and
%\IEEEauthorblockN{Jack Weissenberger}
%\IEEEauthorblockA{\textit{Department of Computer Science} \\
%\textit{Wake Forest University}\\
%Winston-Salem, NC, USA \\
%\email{jack.weissenberger@gmail.com}}
%\and
%\IEEEauthorblockN{Luoping Zhang}
%\IEEEauthorblockA{\textit{Department of Computer Science} \\
%\textit{Wake Forest University}\\
%Winston-Salem, NC, USA \\
%\email{zhanl317@wfu.edu}}
%}

@@

One of the reasons such algorithms are impractical is that they are based on so-called \emph{Arbitrary Precision Approximating} (APA) algorithms, which we describe in detail in \cref{sec:APA}.
In exact arithmetic, these algorithms compute an approximation of the correct result, where the error is polynomial in a nonzero parameter of the algorithm \cite{BLR80}.
That is, in exact arithmetic, the error can be made arbitrarily small.
In floating point arithmetic, however, there is a lower bound on the approximation error that depends on the working precision and properties of the algorithm.
Thus, APA algorithms are often considered to have insufficient accuracy for most applications and have largely been overlooked as practical tools despite their performance potential and ability to outperform their exact counterparts \cite{BB15}.

@@

Our contribution is the use of APA matrix multiplication algorithms to address this problem and accelerate training.
In particular we
\begin{enumerate}
	\item curate a collection of both well-established and recently discovered practical APA algorithms;
	\item extend the framework of \cite{BB15} to generate efficient multithreaded code for all of them, achieving up to 21\% performance improvement over the best parallel classical implementation,
	\item demonstrate the robustness of learning accuracy to approximate matrix multiplications; and
	\item present multithreaded performance improvements of a synthetic MLP of up to 13\% over the use of classical matrix multiplication.
\end{enumerate}

@@

\begin{table}
\centering
\caption{Properties of Arbitrary Precision Approximating Algorithms  \\ Speedup and Error are computed assuming 1 recursive step}
\label{tab:algs}
\begin{tabular}{| c | c c c | c c c |} 
\hline
\textbf{Ref} & \textbf{Dims} & \textbf{Rank} & \textbf{Speedup} & $\mathbf{\sigma}$ & $\mathbf{\varphi}$ & \textbf{Error} \\
\hline
- & $\dims{2,2,2}$ & 8 & - & 1 & 0 & \num{1.2e-7} \\
\hline
\cite{BCRL79} & $\dims{3,2,2}$ & 10 & $20\%$ & 1 & 1 & \num{3.5e-4} \\
\cite{AS13} & $\dims{4,2,2}$ & 13 & $23\%$ & 1 & 2 & \num{4.9e-3} \\
\cite{Smirnov13} & $\dims{3,3,2}$ & 14 & $29\%$ & 1 & 3 & \num{1.9e-2} \\
\cite{Smirnov13} & $\dims{5,2,2}$ & 16 & $25\%$ & 1 & 3 & \num{1.9e-2} \\
\cite{Smirnov13} & $\dims{3,3,3}$ & 20 & $35\%$ & 1 & 6 & \num{1.0e-1} \\
\cite{Schonhage81} & $\dims{3,3,3}$ & 21 & $29\%$ & 1 & 2 & \num{4.9e-3} \\
\cite{Smirnov15} & $\dims{7,2,2}$ & 22 & $27\%$ & 1 & 5 & \num{7.0e-2} \\
\cite{Smirnov16b} & $\dims{4,4,2}$ & 24 & $33\%$ & 1 & 3 & \num{1.9e-2} \\
\cite{Smirnov16a} & $\dims{4,3,3}$ & 27 & $33\%$ & 1 & 3 & \num{1.9e-2} \\
\cite{Smirnov16b} & $\dims{5,5,2}$ & 37 & $35\%$ & 1 & 3 & \num{1.9e-2} \\
\cite{Smirnov14} & $\dims{4,4,4}$ & 46 & $39\%$ & 1 & 3 & \num{1.9e-2} \\
\cite{Smirnov18} & $\dims{5,5,5}$ & 90 & $39\%$ & 1 & 3 & \num{1.9e-2} \\
\hline
\end{tabular}
\end{table}

@@

\begin{figure*}[t!]
%
\subfloat[One thread]{
\centering
\renewcommand{\datafile}{data/matmul_seq.dat}
\renewcommand{\threads}{1}
\input{plots/matmul_perf}
%\caption{Single-threaded matrix multiplication performance}
\label{fig:matmul_seq}}
%
\hfill
\subfloat[Six threads]{
\centering
\renewcommand{\datafile}{data/matmul_par_6.dat}
\renewcommand{\threads}{6}
\input{plots/matmul_perf}
%\caption{Multithreaded matrix multiplication performance (6 threads)}
\label{fig:matmul_par_6}}
%
\hfill
\subfloat[Twelve threads]{
\centering
\renewcommand{\datafile}{data/matmul_par_12.dat}
\renewcommand{\threads}{12}
\input{plots/matmul_perf}
%\caption{Multithreaded matrix multiplication performance (6 threads)}
\label{fig:matmul_par_12}}
%
\caption{Single- and multi-threaded square matrix multiplication performance, relative to $2n^3$ operations}
\label{fig:matmul_perf}
\end{figure*}

>>>>

\documentclass[sigconf,review,anonymous]{acmart}

@@

\author{\IEEEauthorblockN{Grey Ballard}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Wake Forest University}\\
Winston-Salem, NC, USA \\
\email{ballard@wfu.edu}}
\and
\IEEEauthorblockN{Jack Weissenberger}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Wake Forest University}\\
Winston-Salem, NC, USA \\
\email{jack.weissenberger@gmail.com}}
\and
\IEEEauthorblockN{Luoping Zhang}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{Wake Forest University}\\
Winston-Salem, NC, USA \\
\email{zhanl317@wfu.edu}}
}

@@

One of the reasons such algorithms are impractical is that they are based on so-called \emph{Arbitrary Precision Approximating} (APA) algorithms, which we describe in detail in \cref{sec:APA}.
In exact arithmetic, these algorithms compute an approximation of the correct result, where the error is polynomial in a nonzero parameter of the algorithm \cite{BLR80}.
That is, in exact arithmetic, the error can be made arbitrarily small.
In floating point arithmetic, however, there is a lower bound on the approximation error that depends on the working precision and properties of the algorithm.
Thus, APA algorithms are often considered to have insufficient accuracy for most applications and have largely been overlooked as practical tools despite their performance potential and ability to outperform exact algorithms \cite{BB15}.

@@

Our contribution is the use of APA matrix multiplication algorithms to address this problem.
In particular we
\begin{enumerate}
	\item curate a collection of both well-established and recently discovered practical APA algorithms;
	\item extend the framework of \cite{BB15} to generate efficient multithreaded code for all of them, achieving up to 21\% performance improvement over the best parallel classical implementation,
	\item demonstrate the robustness of learning accuracy to approximate matrix multiplications; and
	\item present multithreaded performance improvements of a synthetic MLP of up to 13\% over the use of classical matrix multiplication.
\end{enumerate}

@@

\begin{table}
\centering
\caption{Properties of APA algorithms. Speedup and error are computed assuming 1 recursive step}
\label{tab:algs}
\begin{tabular}{| c | c c c | c c c |} 
\hline
\textbf{Ref} & \textbf{Dims} & \textbf{Rank} & \textbf{Speedup} & $\mathbf{\sigma}$ & $\mathbf{\varphi}$ & \textbf{Error} \\
\hline
- & $\dims{2,2,2}$ & 8 & - & 1 & 0 & \num{1.2e-7} \\
\hline
\cite{BCRL79} & $\dims{3,2,2}$ & 10 & $20\%$ & 1 & 1 & \num{3.5e-4} \\
\cite{AS13} & $\dims{4,2,2}$ & 13 & $23\%$ & 1 & 2 & \num{4.9e-3} \\
\cite{Smirnov13} & $\dims{3,3,2}$ & 14 & $29\%$ & 1 & 3 & \num{1.9e-2} \\
\cite{Smirnov13} & $\dims{5,2,2}$ & 16 & $25\%$ & 1 & 3 & \num{1.9e-2} \\
\cite{Smirnov13} & $\dims{3,3,3}$ & 20 & $35\%$ & 1 & 6 & \num{1.0e-1} \\
\cite{Schonhage81} & $\dims{3,3,3}$ & 21 & $29\%$ & 1 & 2 & \num{4.9e-3} \\
\cite{Smirnov15} & $\dims{7,2,2}$ & 22 & $27\%$ & 1 & 5 & \num{7.0e-2} \\
\cite{Smirnov16b} & $\dims{4,4,2}$ & 24 & $33\%$ & 1 & 3 & \num{1.9e-2} \\
\cite{Smirnov16a} & $\dims{4,3,3}$ & 27 & $33\%$ & 1 & 3 & \num{1.9e-2} \\
\cite{Smirnov16b} & $\dims{5,5,2}$ & 37 & $35\%$ & 1 & 3 & \num{1.9e-2} \\
\cite{Smirnov14} & $\dims{4,4,4}$ & 46 & $39\%$ & 1 & 3 & \num{1.9e-2} \\
\cite{Smirnov18} & $\dims{5,5,5}$ & 90 & $39\%$ & 1 & 3 & \num{1.9e-2} \\
\hline
\end{tabular}
\end{table}

@@

\begin{figure*}[t!]
%
\subfloat[One thread]{
\centering
\renewcommand{\datafile}{data/matmul_seq.dat}
\renewcommand{\threads}{1}
\input{plots/matmul_perf}
%\caption{Single-threaded matrix multiplication performance}
\label{fig:matmul_seq}}
%
\hfill
\subfloat[Six threads]{
\centering
\renewcommand{\datafile}{data/matmul_par_6.dat}
\renewcommand{\threads}{6}
\input{plots/matmul_perf}
%\caption{Multithreaded matrix multiplication performance (6 threads)}
\label{fig:matmul_par_6}}
%
\hfill
\subfloat[Twelve threads]{
\centering
\renewcommand{\datafile}{data/matmul_par_12.dat}
\renewcommand{\threads}{12}
\input{plots/matmul_perf}
%\caption{Multithreaded matrix multiplication performance (6 threads)}
\label{fig:matmul_par_12}}
%
\caption{Single- and multi-threaded square matrix multiplication performance, relative to $2n^3$ operations.  The machine peak with respect to classical algorithms is given as a dotted line.}
\label{fig:matmul_perf}
\end{figure*}
